Hypertension and microbiome

### 2019-08-05 articletwo.Rmd
PREVAL DIAB

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(pwr)
library(gtable)
library(car)
```

@@

```{r my variables}
var.BP <- c("MAP", "SYSTM", "DIASM", "PULSEPRESSURE", "HYPERTENSION")
var.CL.min <- c("BL_AGE", "SEX")
var.CL <- c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB",
            "BL_USE_RX_C03","BL_USE_RX_C07", "BL_USE_RX_C08", "BL_USE_RX_C09")
```

@@

```{r maaslin runs, results = 'hide'} 
 pseq.genus.core <- core(pseq.genus, detection = 0.001, prevalence = 0.01)
 pseq.genus.core.viruses.names <- taxa(pseq.genus.core) %>% mygrep(word = "virus$", complement = TRUE)
 pseq.genus.core.pruned <- prune_taxa(pseq.genus.core.viruses.names, pseq.genus.core)

if (!file.exists("rds/maaslin.rds")) {
     maaslin <- maaslinwrapper(pseq = pseq.genus.core.pruned,
                                    looped = var.BP,
                                    forced = var.CL,
                              taxa = taxa(pseq.genus.core.pruned))

@@

@@

pcoa.df <- merge(pcoa.vectors %>%
                 select(Axis.1, Axis.2, Axis.3),
                 meta(pseq.species), by=0, all = TRUE) %>%
    mutate(STATETREATMENT = factor(paste0(HYPERTENSION, ANYDRUG),
                                   levels = c("11", "10", "01", "00")),
           gMAP = cut(oMAP, seq(60, 160, 15))) %>%
    mutate_at(vars(starts_with("Axis.")), .funs = funs(myscale(.)))

@@

@@

Significant associations between BP-variables and first three PCoA axis

```{r pcoa axis analysis axis 1}
glm.pcoa <- lapply(c2l("Axis.1", "Axis.2", "Axis.3"), function(axis)
    calculateglm(pcoa.df,
                 modelstring = sprintf("%%s ~ %s + %s", axis, paste(var.CL, collapse = "+")),
                 responses = list(model_1 = "oMAP", model_2 = "oSYSTM", model_3 = "oDIASM",
                                          model_4 = "oPULSEPRESSURE", model_5 = "HYPERTENSION"),
                 filterstr = "Axis"))

do.call(rbind, glm.pcoa) %>% 
    select(response, term, estimate, std.error, p.value, conf.low, conf.high) %>%
    mutate_if(is.numeric, round, 3) %>%
    filter(p.value < 0.05) %>%
    arrange(p.value) %>%
    kable

@@

```{r Write figure 2, include=FALSE}
fig2head <- "Principal coordinate analysis for bacterial abundances."
fig2foot <- sprintf("Principal coordinate analysis for bacterial abundances at species level. Bray-Curtis distances are calculated for abundances after compositional transformation. First two axes explain %.1f%% of the variation.", round(100*sum(eig.fracs[c(1,3)]),1))
writeimage(doc, 2, "cache/pcoa-species.png", fig2head, fig2foot)
```

@@

@@

scatter.plot <- function(df,
                         maas.results,
                         rr,
                         abund,
                         covariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB",
                                        "BL_USE_RX_C03", "BL_USE_RX_C07", "BL_USE_RX_C08",
                                        "BL_USE_RX_C09"),
                         prediction = TRUE,
                         plots = TRUE) {
    fo <- sprintf("%s ~ %s + %s", rr, abund, paste0(covariates, collapse="+"))
    model <- lm(as.formula(fo), data = df)

@@

    df.predict <- data.frame(BPV = seq(-5,5,0.1)) %>%
        mutate(SEX = as.factor(0),
               BL_AGE = df %>% pull(BL_AGE) %>% mean,
               BMI = df %>% pull(BMI) %>% mean,
               CURR_SMOKE = as.factor(0),
               Q57X = as.factor(1),
               DIAB = as.factor(0),
               BL_USE_RX_C03 = as.factor(0),
               BL_USE_RX_C07 = as.factor(0),
               BL_USE_RX_C08 = as.factor(0),
               BL_USE_RX_C09 = as.factor(0),
               !!abund := seq(-5,5,0.1)
               )
    prediction <- cbind(df.predict, predict(model, df.predict, interval = 'confidence'))

@@

```{r salt characteristics}
df.genus.salt %>%
    select(-Sample) %>% 
    gather(key,value) %>%
    group_by(key) %>%
    summarise_all(funs(N = n(),
                       Mean = mean(as.numeric(value), na.rm = TRUE),
                       sd = sd(as.numeric(value), na.rm = TRUE),
                       NAs = sum(is.na(value))))  %>%
    mutate_if(is.numeric, round, 3) %>%
    kable()
```

@@

```{r lacto adjusted means}
df.genus.salt.unadjusted <- with(df.genus.salt, data.frame(salt_groups = as.factor(seq(3)),
                                                           SEX = as.factor(0),
                                                           BL_AGE = mean(BL_AGE)))

adjmean.lacto.saltgroup <- predict(lm.lacto.saltgroup,
                               df.genus.salt.unadjusted,
                               interval = "confidence") %>%
    as.data.frame %>%
    dplyr::rename(estimate = fit,
                  conf.low = lwr,
                  conf.high = upr) %>%
    tibble::rownames_to_column("term")
```

@@

```{r forest plot lacto}
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Dietary salt",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(5.38, 5.67), breaks = seq(0, 10, 0.1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.4, 1.5), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)

ggsave(file = "cache/salt-forest.png", plot=gsalt, height=4, width=6, dpi = 300)
```

@@

af.models <- mylmrsqr(df = meta(pseq.species),
         response = "MAP",
         loopcovariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB",
                            "ANYDRUG"))

>>>>

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(pwr)
library(gtable)
library(car)
library(M3C)
```

@@

```{r my variables}
var.BP <- c("MAP", "SYSTM", "DIASM", "PULSEPRESSURE", "HYPERTENSION")
var.CL.min <- c("BL_AGE", "SEX")
var.CL <- c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "PREVAL_DIAB",
            "BL_USE_RX_C03","BL_USE_RX_C07", "BL_USE_RX_C08", "BL_USE_RX_C09")
```

@@

```{r t-sne, eval = FALSE}
sne.df <- pseq.genus %>%
    microbiome::transform(transform = "clr") %>%
    psmelt %>%
    spread(OTU, Abundance)

sne_names <- colnames(sne.df) %>% mygrep(word = "g_")

sne.t <- sne.df %>% select(Sample, HYPERTENSION, sne_names[1:20]) %>%
    mutate(HYPERTENSION = as.numeric(HYPERTENSION)) %>%
    gather(variable, value, -Sample) %>% 
    spread(Sample, value) %>%
    column_to_rownames("variable")

png("sample.png", 800, 800)
tsne(sne.t[-1,], labels = as.factor(sne.t[1,]),
     dotsize = 1)
dev.off()

### CONTINUOUS

source("tsne.R")

sne.t <- sne.df %>% select(Sample, oMAP, sne_names) %>%
    gather(variable, value, -Sample) %>% 
    spread(Sample, value) %>%
    column_to_rownames("variable")

tsne.g <- tsne(sne.t[rownames(sne.t) != "oMAP",],
               labels = as.numeric(sne.t[rownames(sne.t) == "oMAP",]),
               dotsize = 1,
               controlscale = TRUE,
               scale = 2,
               low = "white")

tsne.plot <- tsne.g +
    scale_fill_gradientn(limits = c(-3,3),
                         colours=c("navyblue", "darkmagenta", "darkorange1"),
                         breaks=c(0,200), labels=format(c(0,200)))

ggsave(file = "cache/tsne.png", plot=tsne.plot, height=6, width=9)

```{r maaslin runs, results = 'hide'} 
 pseq.genus.core <- core(pseq.genus, detection = 0.001, prevalence = 0.01)

if (!file.exists("rds/maaslin.rds")) {
     maaslin <- maaslinwrapper(pseq = pseq.genus.core,
                                    looped = var.BP,
                                    forced = var.CL,
                              taxa = taxa(pseq.genus.core))

@@

Hypertension prevalence

```{r test htn prevalence}
dset.alpha %>%
    dplyr::filter(BL_AGE < 60 + 12, BL_AGE > 60 - 12) %>%
    summarize(htn = sum(ifelse(HYPERTENSION == 1, 1, 0)), n = n(), proportion = htn/n) %>%
    kable
```

@@

pcoa.df <- merge(pcoa.vectors,
                 meta(pseq.species), by=0, all = TRUE) %>%
    mutate(STATETREATMENT = factor(paste0(HYPERTENSION, ANYDRUG),
                                   levels = c("11", "10", "01", "00")),
           gMAP = cut(oMAP, seq(60, 160, 15))) %>%
    mutate_at(vars(starts_with("Axis.")), .funs = funs(myscale(.)))

@@

```{r pcoa top axes}
axis_top_labeller <- function(variable, value){
    index <- as.integer(gsub("Axis.", "", value))
    return(sprintf("Axis %i (%.1f%%)", index, 100*eig.fracs[index]))
}

pcoa.plot.top <- ggplot(data=pcoa.df %>% gather(key, Axis, paste0("Axis.", seq(2,43))) %>%
                       mutate(key = factor(key, levels=paste0("Axis.", seq(2,43)))),
                    aes(x=Axis.1, y=Axis, color = HYPERTENSION)) +
    facet_wrap(~key, scales = "free_y", labeller = axis_top_labeller) +
    geom_point(size=0.1, alpha=0.2) +
    xlab("PCoA 1") +
    ylab(NULL) +
    scale_colour_manual(name = "State and treatment",
                        labels = c("Normotensive", "Hypertensive"),
                        breaks=c(0, 1),
                        values = c("blue", "red")) + 
    theme_classic() +
    theme(legend.position = "none")

ggsave(file = "cache/pcoa-species-top-axes.png", plot=pcoa.plot.top,
       height=10, width=10, units = "in", dpi = 300)
```

@@

Significant associations between BP-variables and top PCoA axes

```{r pcoa analysis main axis}
glm.pcoa <- lapply(c2l(paste0("Axis.", seq(1,5))), function(axis)
    calculateglm(pcoa.df,
                 modelstring = sprintf("%%s ~ %s + %s", axis, paste(var.CL, collapse = "+")),
                 responses = list(model_1 = "oMAP", model_2 = "oSYSTM", model_3 = "oDIASM",
                                          model_4 = "oPULSEPRESSURE", model_5 = "HYPERTENSION"),
                 filterstr = "Axis"))

do.call(rbind, glm.pcoa) %>% 
    select(response, term, estimate, std.error, p.value, conf.low, conf.high) %>%
    mutate_if(is.numeric, round, 3) %>%
    arrange(p.value) %>%
    kable

@@

```{r Write figure 2, include=FALSE}
fig2head <- "Principal coordinate analysis for bacterial abundances."
fig2foot <- sprintf("Principal coordinate analysis for bacterial abundances at species level. Bray-Curtis distances are calculated for abundances after compositional transformation. First two axes explain %.1f%% of the variation.", round(100*sum(eig.fracs[1:3]),1))
writeimage(doc, 2, "cache/pcoa-species.png", fig2head, fig2foot)
```

@@

```{r maaslin insignif, include = FALSE, eval = FALSE}
 maaslin %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("[gs]_", "", Feature),
                  beta = Coefficient) %>%
     arrange(Feature) %>%
     select(Variable, Feature, Qval) %>%
     spread(Variable, Qval)
```

@@

scatter.plot <- function(df,
                         maas.results,
                         rr,
                         abund,
                         covariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "PREVAL_DIAB",
                                        "BL_USE_RX_C03", "BL_USE_RX_C07", "BL_USE_RX_C08",
                                        "BL_USE_RX_C09"),
                         prediction = TRUE,
                         plots = TRUE) {
    fo <- sprintf("%s ~ %s + %s", rr, abund, paste0(covariates, collapse="+"))
    model <- lm(as.formula(fo), data = df)

@@

    df.predict <- data.frame(BPV = seq(-5,5,0.1)) %>%
        mutate(SEX = as.factor(0),
               BL_AGE = df %>% pull(BL_AGE) %>% mean,
               BMI = df %>% pull(BMI) %>% mean,
               CURR_SMOKE = as.factor(0),
               Q57X = as.factor(1),
               PREVAL_DIAB = as.factor(0),
               BL_USE_RX_C03 = as.factor(0),
               BL_USE_RX_C07 = as.factor(0),
               BL_USE_RX_C08 = as.factor(0),
               BL_USE_RX_C09 = as.factor(0),
               !!abund := seq(-5,5,0.1)
               )
    prediction <- cbind(df.predict, predict(model, df.predict, interval = 'confidence'))

@@

```{r salt characteristics}
df.genus.salt %>%
    select(-Sample) %>% 
    gather(key,value) %>%
    group_by(key) %>%
    summarise_all(funs(N = n(),
                       Mean = mean(as.numeric(value), na.rm = TRUE),
                       ones = sum(value == 1, na.rm = TRUE),
                       sd = sd(as.numeric(value), na.rm = TRUE),
                       NAs = sum(is.na(value))))  %>%
    mutate_if(is.numeric, round, 3) %>%
    kable()
```

@@

```{r lacto adjusted means}
df.genus.salt.unadjusted <- with(df.genus.salt, data.frame(salt_groups = as.factor(seq(3)),
                                                           SEX = as.factor(1),
                                                           BL_AGE = mean(BL_AGE)))

adjmean.lacto.saltgroup <- predict(lm.lacto.saltgroup,
                               df.genus.salt.unadjusted,
                               interval = "prediction") %>%
    as.data.frame %>%
    dplyr::rename(estimate = fit,
                  conf.low = lwr,
                  conf.high = upr) %>%
    tibble::rownames_to_column("term")
```

@@

```{r forest plot lacto}
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Dietary salt",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(3.0, 8.0), breaks = seq(0, 10, 1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.4, 1.5), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)

ggsave(file = "cache/salt-forest.png", plot=gsalt, height=3, width=5, dpi = 300)
```

@@

af.models <- mylmrsqr(df = meta(pseq.species),
         response = "MAP",
         loopcovariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "PREVAL_DIAB",
                            "ANYDRUG"))
